Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-09-01 interimreport2.Rnw
Draft final interim report 2

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations. We provide preliminary benefit--cost analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:
\begin{itemize}
\item Age is the best predictor of risk of measles infection in multi-variate regression analyses, though some groups such as Pacific people, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is very beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
\end{itemize}

@@

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks. In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand. The additional analyses included in this section are multivariate modelling to account for confounding within the univariate analyses.

@@

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: 1--5 and 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:case}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="Category",cex.lab=2)
legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
@
\end{center}
\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

@@

However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

@@

model2<-update(model1,~.-Age:NZDep)
summary(model2)
anova(model2,test="F")
@
%\begin{figure}[h!]
%\begin{center}
%<<fig=TRUE,echo=FALSE,width = 11>>=
%hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
%@
%\end{center}
%\caption{test}
%\label{fig:test}
%\end{figure}

%\begin{figure}[h!]
%\begin{center}
%<<fig=TRUE,echo=FALSE,width = 11>>=
%plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
%@
%\end{center}
%\caption{Cases per capita for Prioritized Ethnicities. Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity. Points are coloured by Ethnicity.}
%\label{fig:test}
%\end{figure}
%
%\begin{figure}[h!]
%\begin{center}
%<<fig=TRUE,echo=FALSE,width = 11>>=
%res<-predict(model2)
%plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
%@
%\end{center}
%\caption{test}
%\label{fig:test}
%\end{figure}

%<<echo=FALSE,include=FALSE,results=hide>>=
%cor(exp(res),tp$cases)
%cor.test(exp(res),tp$cases)
%@
%
%\begin{figure}[h!]
%\begin{center}
%<<fig=TRUE,echo=FALSE,width = 11>>=
%hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
%@
%\end{center}
%\caption{test}
%\label{fig:test}
%\end{figure}

%\section{Regression analyses results}
%\label{sub:regression_results}
%The results are:

%<<label=regmodel,include=FALSE>>=
%summary(model2)
%@

%<<label=regmodelanova,include=FALSE>>=
%anova(model2,test="F")
%@

@@

The distribution of the cases per category used are in Figure~\fig{fig:histcasecat}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
@
\end{center}
\caption{Histogram of cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width =11, height=11>>=
res<-predict(model3)
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
@
\end{center}
\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
\label{fig:predict}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
@
\end{center}
\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
\label{fig:resid}
\end{figure}
The significance of the different predictor variables can be seen here, in the ANOVA results:

@@

The results of the model suggest that, apart from over-representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).

@@

In later outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young are still most likely to be infections, but older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.

@@

The analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at risk. In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European 6--17 year old children. The results of these analyses are discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR). A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.

@@

A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}. The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.

@@

The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$. This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}. If $R_0 > 1$ an epidemic will occur following an introduction of infection. The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}. The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule. If $R_v < 1$ epidemics are prevented. The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.

@@

To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009. To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}. We were required to compute the generation time for measles to do so. The generation time is the average time an index case infects others after becoming infected. We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}. We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}). The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.

@@

We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand. We estimated this from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}). Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.

We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals. These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).

\subsection{Modelling results}

The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report. The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one. As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious. However, the results of these differences are presented in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).

These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, and we explore the benefit--cost effects of this in Section~\ref{sub:cost_benefit}.

Again The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months. This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak. A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated. However, sub-clinical and underreporting may lower the estimate. The relative contributions of both to our estimates are currently unknown.

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage. Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}). These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).

\subsection{Summary of modelling}
These results suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}). However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary. In Isreal researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached. Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate. 
In summary:
\begin{itemize}
\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
 \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
 \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
\end{itemize}

@@

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}). Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.  The cost was \$256 less if the case were a male. It increased of approximately \$406 per night of hospitalisationand \$64 per year over the time period of 2000 - 2014. The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.

@@

\begin{figure}
     \centering
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(sizes,xlim=c(0,200),col="grey",breaks=3000,xlab="Outbreak size",main="",cex.lab=2)
#summary(sizes)
abline(v=median(sizes),col="orange",lty=3,lwd=10)
abline(v=mean(sizes),col="red",lty=3,lwd=10)
#length(which(sizes>5))/1000
legend('top',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
@
     \caption{test}
     \label{fig:sim}
\end{figure}

>>>>

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations. We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:
\begin{itemize}
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
\end{itemize}

@@

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks. In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand. The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses.

@@

This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: NZDep 1--5 and NZDep 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data.

The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="Category",cex.lab=2)
legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
@
\end{center}
\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
\label{fig:percap}
\end{figure}

@@

However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).

For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model. We also account for differences in population sizes by using an offset term, the log(population size). We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms. Thus, the final model that remained with all significant interaction terms had the following linear predictor:

@@

model2<-update(model1,~.-Age:NZDep)
summary(model2)
anova(model2,test="F")
@

@@

The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
@
\end{center}
\caption{Histogram of cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width =11, height=11>>=
res<-predict(model3)
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
@
\end{center}
\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
\label{fig:predict}
\end{figure}

@@

\begin{figure}[h!]
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
@
\end{center}
\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
\label{fig:resid}
\end{figure}
The significance of the different predictor variables can be seen in the ANOVA results:

@@

The results of the model suggest that, apart from over-representation of some MLA categories discussed above and not included here, age is a strong predictor of being a measles case. Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).

@@

In later outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories. Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.

@@

The regression analyses suggest that age is a particularly strong risk factor for measles. This comes as no surpise to epidemiologists or health care providers. However, our analyses also highlight other groups that are at risk. In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age. Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses. Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR). A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.

The results of these regression analyses are also discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit})

@@

A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}. The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles. Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project. The figure shown (Figure~\ref{fig:naive}) and for all subsequent analyses assume a currently susceptible population in New Zealand of approximately 11\% of the total population.

@@

The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$. This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}. If $R_0 > 1$ an epidemic will occur following an introduction of infection. The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}. The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule. If $R_v < 1$ epidemics are predicted to be prevented (but see below) and infection persistence is prevented. The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.

@@

To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we previously estimated $R_v$ from all the outbreaks in New Zealand since 2009. To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}. We were required to compute the generation time for measles to do so. The generation time is the average time an index case infects others after becoming infected. We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}. We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}). The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.

@@

We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$. We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}). Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.

We then used the proportions of the population that were estimated to require vaccination (Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals. These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).

finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of suscpetible people in New Zealand was now reduced by the numbers vaccinated. Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).

\subsection{Modelling results and discussion}

The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report. The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one. As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious. However, the results of these differences are presented and discussed in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).

These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process. Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we explore the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).

The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months. This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak. A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated. However, sub-clinical and underreporting may lower the estimate. The relative contributions of both to our estimates are currently unknown.

To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:pc}. Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}). These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).

The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}). However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary. Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached. Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.

\subsection{Summary of modelling}
\begin{itemize}
\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
 \item The reproduction number, $R_v$, for measles in the current 2013--2014 outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
 \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
\end{itemize}

@@

Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}). Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.  The cost was \$256 less if the case were a male. It increased of approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014. The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.

@@

\begin{figure}
     \centering
<<fig=TRUE,echo=FALSE,width = 11>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
hist(sizes,xlim=c(0,200),col="grey",breaks=3000,xlab="Outbreak size",main="",cex.lab=2)
#summary(sizes)
abline(v=median(sizes),col="orange",lty=3,lwd=10)
abline(v=mean(sizes),col="red",lty=3,lwd=10)
#length(which(sizes>5))/1000
legend('top',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
@
     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1. The full distribution of results can be seen in Figure~\ref{fig:sim1}}
     \label{fig:sim}
\end{figure}
